Palmer amaranth (\emph{Amaranthus palmeri}) adaptation to agroecossystems

### 2021-08-13 Manuscript/Manuscript.tex
as

% Pandoc citation processing
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
% for Pandoc 2.8 to 2.10.1
\newenvironment{cslreferences}%
  {}%
  {\par}
% For Pandoc 2.11+
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
  % set entry spacing
  \ifnum #2 > 0
  \setlength{\parskip}{#2\baselineskip}
  \fi
 }%
 {}
\usepackage{calc} % for calculating minipage widths
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

@@

Light and temperature are likely the main environment requirements for
Palmer amaranth successful grow and development. Palmer amaranth is
reported with an extended germination period (Jha et al., 2010).
Germination of Palmer amaranth is triggered by 18 C soil temperature
(Keeley et al., 1987), and optimal germination and biomass production
occur at 35/30 C day and night temperatures (Guo and Al-Khatib, 2003).
Palmer amaranth germination is also human-mediated by tillage timings
and preemergence-applied herbicides (Chahal et al., 2021). Water has not
shown to limit Palmer amaranth fitness. Under continuous water stress,
Palmer amaranth survived and produced at least 14000 seeds plant-1
(Chahal et al., 2018). Seeds from Palmer amaranth growing with limited
water conditions were heavier, less dormant, and prompt for germination
(Matzrafi et al., 2021). Also, growing conditions and management can
impact sex dimorphism and flowering pattern in Palmer amaranth (Korres
et al., 2017; Rumpa et al., 2019). The continuous global temperature
warming can impact agriculture and promote niches for Palmer amaranth
invasion/adaptation into new environments. Currently, it is estimated
that the greatest climatic risk of Palmer amaranth establishment are
agronomic crops in Australia and Sub-Sahara Africa (Kistner and
Hatfield, 2018). Temperature is a key factor limiting Palmer amaranth
expansion to cooler geographies (Briscoe Runquist et al., 2019);
however, under future climate change Palmer amaranth is likely to expand
northward into Canada and Northern Europe (Kistner and Hatfield, 2018;
Briscoe Runquist et al., 2019).

Palmer amaranth is already found in agronomic crops of South America
(Larran et al., 2017; Küpper et al., 2017) and Southern Europe (Milani
et al., 2021). In the US, Palmer amaranth is well established at cotton
fields (Garetson et al., 2019) and non-crop areas (Bagavathiannan and
Norsworthy, 2016) in the southern United States but its range is
expanding northward. For example, herbicide resistant Palmer amaranth is
widespread in Nebraska (Oliveira et al., 2021), Michigan (Kohrt et al.,
2017), Connecticut (Aulakh et al., 2021), and potential damaged to
soybean fields in Illinois is reported (Davis et al., 2015). In Iowa, a
study showed that Palmer amaranth is not well adapted compared to
waterhemp (\emph{Amaranthus tuberculatus}) (Baker, 2021). Successful
cases of Palmer amaranth invasion and near to eradication is documented
in Minnesota (Yu et al., 2021). No Palmer amaranth actively growing was
found in Canada; however, Palmer amaranth seeds was detected in sweet
potato slips (Page et al., 2021). Nonetheless, it seems fated the need
to manage Palmer amaranth in agronomic crops throughout multiple
environments in the near future. Strategies on Palmer amaranth
management should encompass the agroecosystem level but not only
attempts to eradicate the weed. Most tactics to manage Palmer amaranth
are based on technology fixes (Scott, 2011), which are short-term (e.g.,
herbicide and/or tillage) rather than long-term weed management. Palmer
amaranth management should be built on minimizing the species ability to
adapt, grow and develop into agroecossystems.

In the southeastern US, early growing Palmer amaranth is well known to
have higher impact on cotton yields than late established plants (MacRae
et al., 2013). In the northern states, Palmer amaranth impact on the
agroecossystem is recent. Studies investigating Palmer amaranth in those
locations is limited due to the plant classification as noxious weed
species (Yu et al., 2021). Nonetheless, the continuous Palmer amaranth
dispersal and potential establishment across northern United States is
concerning and warrant investigations on species morphology in such
environments. Understanding Palmer amaranth biology and growing
strategies under different agroecossystems can enhance our knowledge on
species adaptation. It can also aid on designing proactive and
ecological tactics to limit the species range expansion, reduce its
negative impact, and design resilient and sustainable farming systems
(MacLaren et al., 2020). Therefore, the objective of this study was to
investigate the flowering pattern, biomass production, and height of
Palmer amaranth growing under corn, soybean and fallow at two timings
across five locations in the United States Midwest.

@@

The experiment was conducted in 2018 and 2019 under field conditions at
five locations: Arlington (Washington County, Wisconsin), Clay Center
(Clay County, Nebraska), Grant (Perkins County, Nebraska), Lincoln
(Lancaster County, Nebraska), and Macomb (McDonough County, Illinois).

A glyphosate-resistant soybean cultivar (DSR-1950 R2Y at 296,400 seeds
ha 1), and a corn hybrid were planted at

Monthly mean air temperature and sum precipitation were obtained using
Daymet weather data from June through September across the five
locations in 2018 and 2019 (Correndo et al., 2021) (Figure 1)

The field experimental unit were six adjacent 9.1 m wide (12 rows at
72.2 cm row spacing) by 10.7 m long. Each experimental unit was planted
with corn or soybean (DSR-1950 R2Y at 296,400 seeds ha 1), or under
fallow condition. Palmer amaranth seedlings (potting mix + two
seedlings) were and gently transferring to the ground (6 cm deep and 8
cm wide). Twenty-four plants were equidistantly placed (0.76 m apart)
between rows within each agroecossystems. After a week, one was
eliminated and one was kept. There were two transplant timing: first
(June 1) and second (July 1). There were 24 Palmer amaranth plants in
each experimental unit, with a total of 144 plants for each location.
The study was repeated twice.

@@

Palmer amaranth height and biomass were performed with a linear mixed
model using \emph{lmer} function from ``lme4'' package (Bates et al.,
2015). Plant height and biomass were transformed to meet model
assumption of normality. In the model, agroecosystem (crop, soybean,
fallow) was the fixed effect and year nested with location the random
effects. Analysis of variance was performed with \emph{anova} function
from ``car'' package (Fox and Weisberg, 2018). Marginal means and
compact letter display were estimated with \emph{emmeans} and \emph{cld}
from packages ``emmeans'' and ``multcomp'' (Hothorn et al., 2008).

@@

Palmer amaranth accumulated more biomass when growing in fallow compared
to Palmer amaranth growing in soybean and corn (Figure 3A). At first
cohort time, Palmer amaranth biomass was 75.5 g plant-1, 28.3 g plant-1
and 16.3 g plant-1 in fallow, soybean and corn, respectively. At second
cohort timing, Palmer amaranth produced 62.6 g plant in fallow, followed
by 6.3 g plant in soybean, and 1.4 g plant in corn.

Palmer amaranth height was more uniform across cohort timings, except
when growing in corn (Figure 3B). Palmer amaranth achieve achieve 69.2
cm tall when growing at first cohort timing in bareground, which was not
different to 70.7 cm tall at second cohort timing (P = 0.74). In
addition, no difference in Palmer amaranth height (69.3 cm) was found in
soybean at first cohort compared to fallow (P \textgreater{} 0.75). At
second cohort in soybean, Palmer amaranth was near to 10 cm lower
compared to its first cohort time (P = 0.04). The tallest and smallest
Palmer amaranth plants were found in corn. Palmer amaranth reached 85.2
cm tall at first cohort and 38.2 cm at second cohort timings.

@@

Palmer amaranth growing in corn resulted in a longer flowering pattern
compared to fallow and soybean at first cohort (Figure 2A). Nonetheless,
the 10\% cumulative Palmer amaranth flowering in soybean, fallow and
corn occurred at the end of June. Palmer amaranth reached 10\% flowering
in soybean at doy 180, which was slightly different from fallow (doy
180.9; \emph{P} = 0.01) and corn (doy 181.7; \emph{P} = 0.00). The 50\%
Palmer amaranth cumulative flowering occurred in July. For example,
Palmer amaranth reached 50\% flowering in fallow at doy 193.4, followed
by soybean (doy 194.8), corn (doy 206.6). Similar trend was observed at
90\% Palmer amaranth cumulative flowering. Palmer amaranth growing in
corn reached 90\% flowering at doy 252.6 (early September), which was
37.8 and 32.2 days after Palmer amaranth 90\% flowering in fallow and
soybean, respectively.

Palmer amaranth cumulative flowering at second cohort ranged from mid
July to mid September (Figure 2B). Palmer amaranth growing in fallow
resulted in earlier flowering time compared to soybean and corn. Palmer
amaranth growing in fallow reached 10\%, 50\%, and 90\% flowering time
at day 203.8, 214.4, and 232.2, respectively. Palmer amaranth growing in
soybean reached 10\% flowering at doy 210.9, which was 6 days prior to
corn (\emph{P}-value = 0.00). Similar trend was observed at 50\%
flowering, whereas Palmer amaranth reached 50\% flowering in corn (doy
233.0) 4 days after soybeans (doy 228.9; \emph{P} = 0.00). The 90\%
Palmer amaranth cumulative flowering occurred at same day in corn
(260.9) and soybean (260.5; \emph{P} = 0.66).

@@

Our study showed that Palmer amaranth biomass, height, flowering pattern
and gender varied within agroecossystems and cohort timings. In general,
Palmer amaranth produced more biomass and taller plants when growing at
first cohort rather than second cohort. At first cohort, resources
(e.g., soil nutrients) and conditions (e.g., light) were more timely
available for both species, crop and weed. High biomass and taller
Palmer amaranth plants are likely a weed strategy to compete for light
in between crop rows in absence of canopy. In such condition, Palmer
amaranth showed an extraordinary plasticity to adapt upon the
agroecosystem. This is evident when comparing Palmer amaranth canopy
shape, and its extended flowering pattern when growing into corn
compared to soybean. The Palmer amaranth competition strategy was to
mimic the crop grow and development (Figure 4). These results suggests
that Palmer amaranth can fast evolve life-history traits to adapt into
agroecossystems and cultural practices, including nitrogen fertilization
(Bravo et al., 2018). Our results highlight the Palmer amaranth threat
to agroecossystems as breeding more competitive crop varieties is likely
to select more competitive biotypes (Bravo et al., 2017).

@@

An ecological approach to reduce seed production in Palmer amaranth is
understanding its flowering pattern. Our study suggests that Palmer
amaranth flowering pattern was slightly influenced by cropping systems
and cohort timings. Palmer amaranth growing in fallow and corn resulted
in the overall shortest and longest flowering pattern, respectively.
When growing in soybean, Palmer amaranth flowering pattern was similar
to bareground at first cohort but similar to corn at second cohort
timing. Plant flowering initiation is complex and it depends on
biological and ecological factors (Lang, 1965). We hypothesize that when
growing in high competition (e.g., second cohort), Palmer amaranth
plants tend to initiate flowering early, as well as having an extended
flowering pattern. Early flower initiation is plant strategy when
growing in stress conditions. For example, when growing under water
stress, early flowering in Palmer amaranth resulted in a mismatch
between female and male plants by seven days (Mesgaran et al., 2021). A
mismatch in Palmer amaranth male and female flowering period can
minimize plant outcross, and thus reduce plant seed production. Sex
dimorphism manipulation is considered a potential ecological pest
control (McFarlane et al., 2018; Schliekelman et al., 2005).

The mechanisms of sex-determination in plant species is intriguing and
arouse the curiosity of many scientists, including Darwin (Darwin,
1888). In our study, the gender model performance was decent considering
the biology of plant flowering. A 1:1 male and female sex ratio is a
general evolutionary stable strategy for plant species perpetuation
(Fisher, 1930). However, a slight deviation from 1:1 sex ratio might
occur in some dioecious species. For example, the dioecious
\emph{Halophila stipulacea} is a female-biased plant in its native
habitat, but the naturalized \emph{H. stipulacea} have a 1:1 ratio
(Nguyen et al., 2018). Naturalized of \emph{H. stipulacea} reduced
female-male ratio to expand into its non-native habitat (Nguyen et al.,
2018). Also, biotic and/or abiotic stress can influence plant sex
determination. Palmer amaranth male-to-female ratio was greater under
high plant densities (Korres and Norsworthy, 2017) and after herbicide
application (Rumpa et al., 2019). Our model estimated that late
flowering, heavier and taller Palmer amaranth plants deviated from 1:1
ratio in favor to female plants. It was reported that female Palmer
amaranth plants invested more in height, stem and biomass while male
invested more in leaf area and leaf dry weight under nutrient deficiency
(Korres et al., 2017). Our model also estimated more female plants in
soybean and bareground compared to corn, which might be a result of
plant biomass in each agroecossystem. Our results showed the influence
of life-history and ecological traits on sexual dimorphism in Palmer
amaranth. Sexual dimorphism is documented in other dioecious species
(Barrett and Hough, 2013). For example, stronger female plant
competition and greater male tolerance to herbivory was reported in
\emph{Spinacia oleracea} (Pérez-Llorca and Sánchez Vilas, 2019).
Research on candidate genes for sex determination in \emph{Amaranthus}
species are currently underway but it is far to complete (Montgomery et
al., 2021, 2019). Further studies are also needed to understand the
ecological basis of Palmer amaranth flowering, including the plant
behavior under climate change.

Our study demonstrated the short-term Palmer amaranth plasticity to grow
and develop into cropping-systems. Palmer amaranth management priority
should focus on minimizing Palmer amaranth dispersal into new
geographies. Flower manipulation is a promising ecological Palmer
amaranth but it is not well understood. Nonetheless, early-season
management programs would have a large negative effect in Palmer
amaranth growth and development. Tactics that promote early-season crop
advantage against Palmer amaranth, including early crop planting, crop
rotation (Oliveira et al., 2021), plant width, preemergence applied
herbicide (Sanctis et al., 2021), and crop residue (e.g.~cover crops)
would minimize the negative impact of Palmer amaranth in
agroecossystems.

@@

@@

{\centering \includegraphics[width=160mm,height=100mm]{../data analysis/weather/Figure 1}

@@

\caption{Mean average temperature (C) and montly sum precipitation (mm) at Arlington, WI, Clay Center, NE, Grant, NE, Lincoln, NE and Macomb, IL}\label{fig:Figure-1}
\end{figure}

@@

{\centering \includegraphics[width=130mm,height=130mm]{../data analysis/figures/Figure 2}

@@

\caption{Cumulative flowering of Palmer amaranth at first and second transplant timing (A) and day of year of 10, 50, and 90 cumulative flowering at first and second transplant timing (B)}\label{fig:Figure-2}
\end{figure}

@@

{\centering \includegraphics[width=130mm,height=70mm]{../data analysis/figures/Figure 3}

@@

\caption{Palmer amaranth biomass (A) and height (B) growing in corn, fallow, and soybean across Arlington, Clay Center, Grant, Lincoln, and Macomb}\label{fig:Figure-3}
\end{figure}

{\centering \includegraphics[width=70mm,height=70mm]{../data analysis/figures/image}

\caption{Harvest Palmer amaranth plants at 40 days after first transplant timing. From left to right, Palmer amaranth growing in fallow, soybean and corn in Arlington, Wisconsin}\label{fig:Figure-4}
\end{figure}

>>>>

% Pandoc citation processing
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
% for Pandoc 2.8 to 2.10.1
\newenvironment{cslreferences}%
  {}%
  {\par}
% For Pandoc 2.11+
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
  % set entry spacing
  \ifnum #2 > 0
  \setlength{\parskip}{#2\baselineskip}
  \fi
 }%
 {}
\usepackage{calc} % for calculating minipage widths
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

@@

Light and temperature are likely the main environment requirements for
Palmer amaranth successful grow and development. Palmer amaranth is
reported with an extended germination period (Jha et al., 2010).
Germination of Palmer amaranth is triggered by 18 C soil temperature
(Keeley et al., 1987), and optimal germination and biomass production
occur at 35/30 C day and night temperatures (Guo and Al-Khatib, 2003).
Palmer amaranth germination is also human-mediated by tillage timings
and preemergence-applied herbicides (Chahal et al., 2021). Water has not
shown to limit Palmer amaranth fitness. Under continuous water stress,
Palmer amaranth survived and produced at least 14000 seeds plant-1
(Chahal et al., 2018). Seeds from Palmer amaranth growing with limited
water conditions were heavier, less dormant, and prompt for germination
(Matzrafi et al., 2021). Growing conditions and management practices
also influence Palmer amaranth sex dimorphism and flowering pattern
(Korres et al., 2017; Rumpa et al., 2019). The continuous global
temperature warming can impact agriculture and promote niches for Palmer
amaranth invasion/adaptation into new environments. Currently, it is
estimated that the greatest climatic risk of Palmer amaranth
establishment are agronomic crops in Australia and Sub-Sahara Africa
(Kistner and Hatfield, 2018). Temperature is a key factor limiting
Palmer amaranth expansion to cooler geographies (Briscoe Runquist et
al., 2019); however, under future climate change Palmer amaranth is
likely to expand northward into Canada and Northern Europe (Kistner and
Hatfield, 2018; Briscoe Runquist et al., 2019).

Palmer amaranth is already found in agronomic crops of South America
(Larran et al., 2017; Küpper et al., 2017) and Southern Europe (Milani
et al., 2021). In the US, Palmer amaranth is well established at cotton
fields (Garetson et al., 2019) and non-crop areas (Bagavathiannan and
Norsworthy, 2016) in the southern United States but its range is
expanding northward. For example, herbicide resistant Palmer amaranth is
widespread in Nebraska (Oliveira et al., 2021), Michigan (Kohrt et al.,
2017), Connecticut (Aulakh et al., 2021), and potential damaged to
soybean fields in Illinois is reported (Davis et al., 2015). In Iowa, a
study showed that Palmer amaranth is still not well adapted compared to
waterhemp (\emph{Amaranthus tuberculatus}) (Baker, 2021). Successful
cases of Palmer amaranth invasion and near to eradication is documented
in Minnesota (Yu et al., 2021). No Palmer amaranth actively growing was
found in Canada; however, Palmer amaranth seeds was detected in sweet
potato slips (Page et al., 2021). Nonetheless, it seems fated the need
to manage Palmer amaranth in agronomic crops throughout multiple
environments in the near future. Strategies on Palmer amaranth
management should encompass the agroecosystem level but not only
attempts to eradicate the weed. Most tactics to manage Palmer amaranth
are based on technology fixes (Scott, 2011), which are short-term (e.g.,
herbicide and/or tillage) rather than long-term weed management. Palmer
amaranth management should be built on minimizing the species ability to
adapt, grow and develop into agroecossystems.

In the southeastern US, early growing Palmer amaranth is well known to
have higher impact on cotton yields than late established plants (MacRae
et al., 2013). In the northern states, Palmer amaranth impact on the
agroecossystem is recent. Studies investigating Palmer amaranth in those
locations is limited due to the plant classification as noxious weed
species (Yu et al., 2021). Nonetheless, the continuous Palmer amaranth
dispersal and potential establishment across northern United States is
concerning and warrant investigations on species morphology in such
environments. Understanding Palmer amaranth biology and growing
strategies under different agroecossystems can enhance our knowledge on
species adaptation. It can also aid on designing proactive and
ecological tactics to limit the species range expansion, reduce its
negative impact, and design resilient and sustainable farming systems
(MacLaren et al., 2020). Therefore, the objective of this study was to
investigate the flowering pattern, gender, biomass production, and
height of Palmer amaranth growing under corn, soybean and bareground at
two timings across five locations in the United States Midwest.

@@

The experiment was conducted in 2018 and 2019 under field conditions at
five locations: Arlington, WI (43°18'N, 89°29'W), Clay Center, NE ('N,
'W), Grant (('N, 'W)), Lincoln (('N, 'W)), and Macomb (('N, 'W)).

Fields were conventionally tilled and disked prior to planting. Corn
hybrid and soybean varieties were planted in 76-cm row spacing (Table
1). Monthly mean air temperature and sum precipitation were obtained
using Daymet weather data from June through September across the five
locations in 2018 and 2019 (Correndo et al., 2021) (Figure 1)

\begin{table}[!h]

\caption{\label{tab:unnamed-chunk-2}Field study attributes }
\centering
\resizebox{\linewidth}{!}{
\fontsize{12}{14}\selectfont
\begin{tabular}[t]{lllllll}
\toprule
Attributes &  & Arlington, WI & Clay Center, NE & Grant, NE & Lincoln, NE & Macomb, IL\\
\midrule
\cellcolor{gray!6}{Bareground} & \cellcolor{gray!6}{} & \cellcolor{gray!6}{} & \cellcolor{gray!6}{} & \cellcolor{gray!6}{} & \cellcolor{gray!6}{} & \cellcolor{gray!6}{}\\

 & Weed control & glyphosate & saflufenacil + imazethapyr + pyroxasulfone &  &  & \\

\cellcolor{gray!6}{Corn} & \cellcolor{gray!6}{} & \cellcolor{gray!6}{} & \cellcolor{gray!6}{} & \cellcolor{gray!6}{} & \cellcolor{gray!6}{} & \cellcolor{gray!6}{}\\

 & Hybrid &  & DKC60-67 &  &  & \\

\cellcolor{gray!6}{} & \cellcolor{gray!6}{Seeding rate} & \cellcolor{gray!6}{} & \cellcolor{gray!6}{86487} & \cellcolor{gray!6}{} & \cellcolor{gray!6}{} & \cellcolor{gray!6}{}\\

 & Weed control & glyphosate / S-metolachor & S-metolachlor + trazine + mesotrione, + bicyclopyrone &  &  & \\

\cellcolor{gray!6}{Soybean} & \cellcolor{gray!6}{} & \cellcolor{gray!6}{} & \cellcolor{gray!6}{} & \cellcolor{gray!6}{} & \cellcolor{gray!6}{} & \cellcolor{gray!6}{}\\

 & Variety & DSR-1950 & AG21X8 &  &  & \\

\cellcolor{gray!6}{} & \cellcolor{gray!6}{Seeding rate} & \cellcolor{gray!6}{296400} & \cellcolor{gray!6}{321237} & \cellcolor{gray!6}{} & \cellcolor{gray!6}{} & \cellcolor{gray!6}{}\\

 & Weed control & glyphosate / S-metolachor & saflufenacil + imazethapyr + pyroxasulfone &  &  & \\

\cellcolor{gray!6}{Planting day} & \cellcolor{gray!6}{} & \cellcolor{gray!6}{May 10 to 20} & \cellcolor{gray!6}{May 10 to 14} & \cellcolor{gray!6}{} & \cellcolor{gray!6}{} & \cellcolor{gray!6}{}\\

Soil &  &  &  &  &  & \\

\cellcolor{gray!6}{} & \cellcolor{gray!6}{Type} & \cellcolor{gray!6}{} & \cellcolor{gray!6}{Crete Silt Loam} & \cellcolor{gray!6}{} & \cellcolor{gray!6}{} & \cellcolor{gray!6}{}\\

 & Ratio (sand-clay-silt) &  & 58-25-6.5 &  &  & \\

\cellcolor{gray!6}{} & \cellcolor{gray!6}{pH} & \cellcolor{gray!6}{6.6} & \cellcolor{gray!6}{6.5} & \cellcolor{gray!6}{} & \cellcolor{gray!6}{} & \cellcolor{gray!6}{}\\

 & Organic matter (\%) &  & 3 &  &  & \\
\bottomrule
\end{tabular}}
\end{table}

The field experimental unit were six adjacent 9.1 m wide (12 rows at
72.2 cm row spacing) by 10.7 m long. Each experimental unit was planted
with corn or soybean, or under bareground condition. Palmer amaranth
seedlings (potting mix + two seedlings) were and gently transferring to
the ground (6 cm deep and 8 cm wide). Twenty-four plants were
equidistantly placed (0.76 m apart) between rows within each
agroecossystems. After a week, one was eliminated and one was kept.
There were two transplant timing: first (June 1) and second (July 1).
There were 24 Palmer amaranth plants in each experimental unit, with a
total of 144 plants for each location. The study was repeated twice.

@@

Palmer amaranth height and biomass were performed with a linear mixed
model using \emph{lmer} function from ``lme4'' package (Bates et al.,
2015). Plant height and biomass were transformed to meet model
assumption of normality. In the model, agroecosystem (crop, soybean,
bareground) was the fixed effect and year nested with location the
random effects. Analysis of variance was performed with \emph{anova}
function from ``car'' package (Fox and Weisberg, 2018). Marginal means
and compact letter display were estimated with \emph{emmeans} and
\emph{cld} from packages ``emmeans'' and ``multcomp'' (Hothorn et al.,
2008).

@@

Palmer amaranth accumulated more biomass when growing in bareground
compared to Palmer amaranth growing in soybean and corn (figure 2A). At
first cohort time, Palmer amaranth biomass was 75.5 g plant-1, 28.3 g
plant-1 and 16.3 g plant-1 in bareground, soybean and corn,
respectively. At second cohort timing, Palmer amaranth produced 62.6 g
plant in bareground, followed by 6.3 g plant in soybean, and 1.4 g plant
in corn.

Palmer amaranth height was more uniform across cohort timings, except
when growing in corn (figure 2B). Palmer amaranth achieved 69.2 cm tall
when growing at first cohort timing in bareground, which was not
different to 70.7 cm tall at second cohort timing (P = 0.74). In
addition, no difference in Palmer amaranth height (69.3 cm) was found in
soybean at first cohort compared to bareground (P \textgreater{} 0.75).
At second cohort in soybean, Palmer amaranth was near to 10 cm lower
compared to its first cohort time (P = 0.04). The tallest and smallest
Palmer amaranth plants were found in corn. Palmer amaranth reached 85.2
cm tall at first cohort and 38.2 cm at second cohort timings.

\vspace{0.4cm}

@@

Palmer amaranth growing in corn resulted in a longer flowering pattern
compared to bareground and soybean at first cohort (figure3A).
Nonetheless, the 10\% cumulative Palmer amaranth flowering in soybean,
bareground and corn occurred at the end of June. Palmer amaranth reached
10\% flowering in soybean at doy 180, which was slightly different from
bareground (doy 180.9; \emph{P} = 0.01) and corn (doy 181.7; \emph{P} =
0.00). The 50\% Palmer amaranth cumulative flowering occurred in July.
For example, Palmer amaranth reached 50\% flowering in bareground at doy
193.4, followed by soybean (doy 194.8), corn (doy 206.6). Similar trend
was observed at 90\% Palmer amaranth cumulative flowering. Palmer
amaranth growing in corn reached 90\% flowering at doy 252.6 (early
September), which was 37.8 and 32.2 days after Palmer amaranth 90\%
flowering in bareground and soybean, respectively.

Palmer amaranth cumulative flowering at second cohort ranged from mid
July to mid September (figure3B). Palmer amaranth growing in bareground
resulted in earlier flowering time compared to soybean and corn. Palmer
amaranth growing in bareground reached 10\%, 50\%, and 90\% flowering
time at day 203.8, 214.4, and 232.2, respectively. Palmer amaranth
growing in soybean reached 10\% flowering at doy 210.9, which was 6 days
prior to corn (\emph{P}-value = 0.00). Similar trend was observed at
50\% flowering, whereas Palmer amaranth reached 50\% flowering in corn
(doy 233.0) 4 days after soybeans (doy 228.9; \emph{P} = 0.00). The 90\%
Palmer amaranth cumulative flowering occurred at same day in corn
(260.9) and soybean (260.5; \emph{P} = 0.66).

\vspace{0.4cm}

@@

Our study showed that Palmer amaranth biomass, height, flowering pattern
and gender varied within agroecossystems and cohort timings. In general,
Palmer amaranth produced more biomass and taller plants when growing at
first cohort rather than second cohort. At first cohort, resources
(e.g., soil nutrients) and conditions (e.g., light) were more timely
available for both species, crop and weed. High biomass and taller
Palmer amaranth plants are likely a weed strategy to compete for light
in between crop rows in absence of canopy. In such condition, Palmer
amaranth showed an extraordinary plasticity to adapt upon the
agroecosystem. This is evident when comparing Palmer amaranth canopy
shape, and its extended flowering pattern when growing into corn
compared to soybean. The Palmer amaranth competition strategy was to
mimic the crop grow and development (Figure 4). These results suggests
that Palmer amaranth can fast evolve life-history traits to adapt into
agroecossystems and cultural practices, including nitrogen fertilization
(Bravo et al., 2018). Our results highlight the Palmer amaranth as a
threat to agroecossystems as breeding more competitive crop varieties is
likely to select more competitive biotypes (Bravo et al., 2017).

@@

An ecological approach to reduce seed production in Palmer amaranth is
understanding its flowering pattern. Our study suggests that Palmer
amaranth flowering pattern was slightly influenced by cropping systems
and cohort timings. Palmer amaranth growing in bareground and corn
resulted in the overall shortest and longest flowering pattern,
respectively. When growing in soybean, Palmer amaranth flowering pattern
was similar to bareground at first cohort but similar to corn at second
cohort timing. Plant flowering initiation is complex and it depends on
biological and ecological factors (Lang, 1965). We hypothesize that when
growing in high competition (e.g., second cohort), Palmer amaranth
plants tend to initiate flowering early, as well as having an extended
flowering pattern. Early flower initiation is plant strategy when
growing in stress conditions. For example, when growing under water
stress, early flowering in Palmer amaranth resulted in a mismatch
between female and male plants by seven days (Mesgaran et al., 2021). A
mismatch in Palmer amaranth male and female flowering period can
minimize plant outcross, and thus reduce plant seed production. Sex
dimorphism manipulation is considered a potential ecological pest
control (McFarlane et al., 2018; Schliekelman et al., 2005).

The mechanisms of sex-determination in plant species is intriguing and
arouse the curiosity of many scientists, including Darwin (Darwin,
1888). In our study, the gender model performance was decent considering
the biology of plant flowering. A 1:1 male and female sex ratio is a
general evolutionary stable strategy for plant species perpetuation
(Fisher, 1930). However, a slight deviation from 1:1 sex ratio might
occur in some dioecious species. For example, the dioecious
\emph{Halophila stipulacea} is a female-biased plant in its native
habitat, but the naturalized \emph{H. stipulacea} have a 1:1 ratio
(Nguyen et al., 2018). Naturalized of \emph{H. stipulacea} reduced
female-male ratio to expand into its non-native habitat (Nguyen et al.,
2018). Also, biotic and/or abiotic stress can influence plant sex
determination. Palmer amaranth male-to-female ratio was greater under
high plant densities (Korres and Norsworthy, 2017) and after herbicide
application (Rumpa et al., 2019). Our model estimated that late
flowering, heavier and taller Palmer amaranth plants deviated from 1:1
ratio in favor to female plants. It was reported that female Palmer
amaranth plants invested more in height, stem and biomass while male
invested more in leaf area and leaf dry weight under nutrient deficiency
(Korres et al., 2017). Our model also estimated more female plants in
soybean and bareground compared to corn, which might linked to plant
competition strategy in each agroecossystem. Our results showed the
influence of life-history and ecological traits on sexual dimorphism in
Palmer amaranth. Sexual dimorphism is documented in other dioecious
species (Barrett and Hough, 2013). For example, stronger female plant
competition and greater male tolerance to herbivory was reported in
\emph{Spinacia oleracea} (Pérez-Llorca and Sánchez Vilas, 2019).
Research on candidate genes for sex determination in \emph{Amaranthus}
species are currently underway but it is far to complete (Montgomery et
al., 2021, 2019). Further studies are also needed to understand the
ecological basis of Palmer amaranth flowering, including the plant
behavior under climate change.

Our study demonstrated the short-term Palmer amaranth plasticity to grow
and develop into cropping-systems. Is likely that Palmer amaranth range
will continue to expanding to new geographies. Therefore, preventive
management is a priority to minimizing Palmer amaranth dispersal.
Reactive management should focus on early-season management programs,
which would have a large negative effect on Palmer amaranth growth and
development. Tactics that promote early-season crop advantage against
Palmer amaranth, including early crop planting, crop rotation (Oliveira
et al., 2021), plant width, preemergence applied herbicide (Sanctis et
al., 2021), and crop residue (e.g.~cover crops) would minimize the
negative impact of Palmer amaranth in agroecossystems. The
aggressiveness and differential Palmer amaranth adaptation to
agroecosystem is striking and require national efforts to minimize the
species impact on economy and sustainability.

@@

Flower manipulation is a promising ecological Palmer amaranth but it is
not well understood.

@@

{\centering \includegraphics[width=150mm,height=90mm]{../data analysis/figures/Figure 3}

@@

\caption{Palmer amaranth biomass (A) and height (B) growing in corn, bareground, and soybean across Arlington, WI, Clay Center, NE, Grant, NE, Lincoln, NE and Macomb, IL}\label{fig:Figure-2}
\end{figure}

@@

{\centering \includegraphics[width=90mm,height=90mm]{../data analysis/figures/image}

@@

\caption{Harvest Palmer amaranth plants at 40 days after first transplant timing. From left to right, Palmer amaranth growing in bareground, soybean and corn in Arlington, Wisconsin}\label{fig:Figure-3}
\end{figure}

@@

{\centering \includegraphics[width=150mm,height=150mm]{../data analysis/figures/Figure 2}

@@

\caption{Cumulative flowering of Palmer amaranth at first and second transplant timing (A) and day of year of 10, 50, and 90 cumulative flowering at first and second transplant timing (B)}\label{fig:Figure-4}
\end{figure}
